Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-10-02 paper/paper.tex
edited abstract and small edits in pill section of paper Tested everything on wentworth.

\begin{abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system usually exhibits robust polar
  oscillations in a variety of cell shapes, new experiments have shown
  that when the cells are mechanically deformed into wide, flattened
  out, irregular shapes, the robust regularity of this oscillatory
  behavoir is no longer seen. Here we study widely used stochastic and
  deterministic models of Min-system simulation within these new cell
  shapes.  We find that the deterministic model is largely inadequate
  in reproducing the qualitatively irregular behavoir, while the
  stochastic model, based on the same differential equations, is able
  to accurately predict what has been seen.  We also find that it is
  indeed the irregularity of the cell shape that is responsible for
  the irregular oscillation behavoir, as opposed to the newly deformed
  cells' increased width and altered aspect ratio.
\end{abstract}

@@

Figure~\ref{image-p} begins about 300 seconds into the simulation and
shows one period of oscillation.  At $t=0$ there is a high
concentration of MinD:ATP that has accumulated on the membrane at the
bottom of the cell. An important aspect of Huang's model is that the
MinD is attracted to and sticks to the membrane nonlinearly: as it
accumulates there it begins to `recruit' other MinD that is diffusing
in the cytoplasm nearby, causing peaks in concentration to build up on
the walls.  Meanwhile, the MinE is creeping downward as is bound to
the membrane-bound MinD, forming the MinD:MinE:ATP complex, then
breaks it apart and diffuses downward a bit more before it again
reacts with membrane-bound MinD.  This process can seen in the form of
the well-known ``MinE rings'' (actually, MinE bound to MinD on the
membrane).  These rings are visible in the deterministic model plots
as a green band on the walls from 0 seconds to 4 seconds (and later
from 15 to 23 seconds).  The appearence of these rings in the
deterministic model along side their less obvious appearence in the
stochastic model highlights an advantage of the deterministic model:
the idealization of deterministic data allows one to see patterns in
the averaged behavior that might otherwise be missed.  In the
stochastic model the MinE still exhibits higher concentrations in the
same regions throughout the process, but what would ideally be a
``ring'' pattern is instead an asymmetric collection of maxima, which
would become a ring after phase-locked averaging.

@@

It is clear from Fig.~\ref{image-p} that the stochastic simulation
results in a protein distribution that is spatially regular than the
deterministic model predicts.  In fact, when viewed as a movie (see
supplementary material), the stochastic model exhibits a ``starry
night'' effect, as recruitment leads to clusters of MinD forming on
the wall and then subsequently dissipating.  This naturally leads to
irregularity in the temporal periodicity of the system, which we
quantify by examining the temporal correlation function of the total
MinD found in two opposite polar regions.
%
This correlation function, which is displayed in Fig.~\ref{corr-pill},
is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom sections of a cell that has been
divided into thirds along its long axis.  These plots help us in
studying the periods and the regularity and stability of oscillations.
These curves are well fit with a simple exponential decoherence model,
given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 7 periods and in the case of the determinisitic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.

>>>>

\begin{abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system usually exhibits robust polar
  oscillations in a variety of cell shapes, new experiments have shown
  that when the cells are mechanically deformed into wide, flattened
  out, irregular shapes, the robust regularity of this oscillatory
  behavoir is no longer seen. Here we study widely used stochastic and
  deterministic models of Min-system simulation within these new cell
  shapes.  We find that the deterministic model is largely inadequate
  in reproducing the qualitatively irregular behavoir, while the
  stochastic model, based on the same differential equations, is able
  to accurately predict what has been seen.  We also find that it is
  the newly deformed cells' increased width and altered aspect ratio
  that is responsible for the irregular oscillation behavoir, as
  opposed to the irregularity of the cell shape.
\end{abstract}

@@

Figure~\ref{image-p} begins about 300 seconds into the simulation and
shows one period of oscillation.  At $t=0$ there is a high
concentration of MinD:ATP that has accumulated on the membrane at the
bottom of the cell. An important aspect of Huang's model is that the
MinD is attracted to and sticks to the membrane nonlinearly: as it
accumulates there it begins to `recruit' other MinD that is diffusing
in the cytoplasm nearby, causing peaks in concentration to build up on
the walls.  Meanwhile, the MinE creeps downward as it reacts with the
membrane-bound MinD, forming the MinD:MinE:ATP complex, then breaks it
apart and diffuses downward a bit more before it again reacts with
membrane-bound MinD.  This process can be seen in the form of the
well-known ``MinE rings'' (actually, MinE bound to MinD on the
membrane).  These rings are visible in the deterministic model plots
as a green band on the walls from 0 seconds to 4 seconds (and later
from 15 to 23 seconds).  The appearence of these rings in the
deterministic model along side their less obvious appearence in the
stochastic model highlights an advantage of the deterministic model:
the idealization of deterministic data allows one to see patterns in
the averaged behavior that might otherwise be missed.  In the
stochastic model the MinE still exhibits higher concentrations in the
same regions throughout the process, but what would ideally be a
``ring'' pattern is instead an asymmetric collection of maxima, which
would become a ring after phase-locked averaging.

@@

It is clear from Fig.~\ref{image-p} that the stochastic simulation
results in a protein distribution that is less spatially regular than
the deterministic model predicts.  In fact, when viewed as a movie
(see supplementary material), the stochastic model exhibits a ``starry
night'' effect, as recruitment leads to clusters of MinD forming on
the wall and then subsequently dissipating.  This naturally leads to
irregularity in the temporal periodicity of the system, which we
quantify by examining the temporal correlation function of the total
MinD found in two opposite polar regions.
%
This correlation function, which is displayed in Fig.~\ref{corr-pill},
is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom sections of a cell that has been
divided into thirds along its long axis.  These plots help us in
studying the periods and the regularity and stability of oscillations.
These curves are well fit with a simple exponential decoherence model,
given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods and in the case of the determinisitic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.
